---
title: "FINAL INDIVIDUAL REPORT"
author: "Siya Aneja"
output: html_document
format:
  html:
    theme: journal
    toc: true
    code-fold: true
execute:
  echo: true
  message: false
  warning: false
  cache: true
  freeze: auto
editor: 
  markdown: 
    wrap: 72
---

## 1. Introduction

Obesity is a major public health concern in the United States, with
prevalence varying across states, age groups, and time. Understanding
whether observed differences represent true population trends or data
instability is crucial for interpreting survey-based health data and
informing policy decisions. This project uses data from the Behavioral
Risk Factor Surveillance System (BRFSS) to analyze obesity prevalence
across U.S. states. The analysis focuses on identifying overall
patterns, state-level deviations, and explaining why certain
states—specifically Florida and West Virginia—appear to behave
atypically.

## 2. GROUP OVERALL QUESTION

The group project investigates the following Overall Question (OQ):

> How do obesity rates evolve across age groups and geographic regions
> in the United States, and which states exhibit atypical patterns?

This question frames obesity as a population-level phenomenon,
emphasizing how prevalence varies across place (states, regions) and
time or age, rather than focusing solely on individual behavior.
Answering this question requires not only identifying broad national and
regional patterns, but also understanding why some states deviate from
those patterns.

## **3.** Individual Sub-Question (SQ)

During the group exploratory analysis, most states displayed smooth,
gradual obesity trends, consistent with expectations for a slow-moving
public health outcome. However, two states—Florida and West
Virginia—stood out as apparent anomalies:

-   Florida showed sharp year-to-year fluctuations (“zig-zag” behavior)

-   West Virginia exhibited an unusually steep upward trend.

These observations motivated the following Specific Question (SQ)
assigned to this individual report:

> What caused obesity rates in Florida and West Virginia to act “weird”?

This SQ plays a critical supporting role in answering the OQ. If these
atypical patterns represent true population dynamics, they would suggest
meaningful geographic exceptions that must be incorporated into the
group’s conclusions. Conversely, if these patterns are driven by data
quality or measurement artifacts, they should not be interpreted as
substantive deviations.

Thus, resolving this SQ is essential to determining whether Florida and
West Virginia represent:

-   genuine geographic outliers in obesity evolution, or

-   statistical anomalies that should be interpreted with caution.

By diagnosing the source of these “weird” trends, this analysis
strengthens the credibility and interpretability of the group’s overall
findings.

## 4. Data Source and Acquisition

We use data from the **Behavioral Risk Factor Surveillance System
(BRFSS)**, administered by the Centers for Disease Control and
Prevention (CDC). The BRFSS collects self-reported health indicators
from U.S. adults, including height and weight, which are used to
estimate obesity prevalence.

### Step 1: Load Required Libraries

```{r}
library(tidyverse)
library(janitor)
library(broom)
```

### Step 2: Download the Data

```{r}
file.copy(
  from = "/Users/siyaaneja/Desktop/Nutrition__Physical_Activity__and_Obesity_-_Behavioral_Risk_Factor_Surveillance_System-2.csv",
  to   = "data/brfss_obesity.csv"
)

```

### Step 3: Inspect the Raw Data

```{r}
brfss_raw <- read_csv(
"data/brfss_obesity.csv",
show_col_types = FALSE
)
nrow(brfss_raw)
glimpse(brfss_raw)

```

## 4. Data Cleaning and Preparation

We restrict the dataset to:

-   Overall adult obesity prevalence

-   State-level observations

-   Non-missing numeric estimates

```{r}
brfss_clean <- brfss_raw %>%
  filter(
    Topic == "Obesity / Weight Status",
    Question == "Percent of adults aged 18 years and older who have obesity"
  ) %>%
  select(
    state   = LocationDesc,
    year    = YearStart,
    obesity = Data_Value
  ) %>%
  filter(!is.na(obesity))

```

```{r}
nrow(brfss_clean)
summary(brfss_clean$obesity)
length(unique(brfss_clean$state))

```

## 5. Diagnostic Assessment: Can Year-to-Year Change Be Estimated?

Before attempting to explain Florida’s and West Virginia’s unusual
patterns, it is necessary to assess whether the data structure supports
formal year-to-year trend estimation at the state level.

```{r}
brfss_clean %>%
group_by(state) %>%
summarise(n_years = n_distinct(year)) %>%
arrange(n_years)

```

***Interpretation***

Most U.S. states contain between 12 and 14 years of obesity
observations, while some territories have fewer years of coverage. This
indicates that limited trend analysis is possible, but the temporal
structure is uneven across states. Importantly, the BRFSS is a repeated
cross-sectional survey rather than a true panel. As a result, changes
across years may reflect:

-   survey redesigns,

-   sampling variability, or

-   changes in weighting procedures, rather than true population-level
    shifts.

This diagnostic step motivates cautious interpretation of state-level
trends, particularly when large year-to-year fluctuations are observed.

## 6. Florida Analysis: Measuring Instability in Obesity Estimates

Florida appeared unusual in the group analysis due to sharp
fluctuations\
in obesity prevalence across visualizations. Given the lack of\
longitudinal structure, these fluctuations are evaluated using\
**within-state variability** rather than time trends.

```{r}
state_variability <- brfss_clean %>%
group_by(state) %>%
summarise(
obesity_sd = sd(obesity, na.rm = TRUE)
) %>%
arrange(desc(obesity_sd))

```

```{r}
state_variability %>%
mutate(rank = row_number()) %>%
filter(state == "Florida")

```

***Interpretation***

Florida ranks among the states with the highest standard deviation in
obesity prevalence. Because obesity is a slow-moving population health
outcome, such high variability is unlikely to reflect true behavioral or
biological change.

Instead, this instability is consistent with:

-   sampling variability,

-   survey weighting effects, or

-   measurement noise in large, heterogeneous populations.

Thus, Florida’s apparent “zig-zag” behavior is best interpreted as a
statistical artifact, not a meaningful deviation from national obesity
patterns.

## 7. West Virginia Analysis: Identifying Structural Obesity Risk

Unlike Florida, West Virginia appeared unusual due to its consistently
high obesity levels. To evaluate whether this represents a genuine
structural pattern, we estimate the **direction and steepness** of
obesity prevalence across the available ordering variable.

```{r}
state_slopes <- brfss_clean %>%
group_by(state) %>%
summarise(
slope = coef(lm(obesity ~ as.numeric(year)))[2]
) %>%
arrange(desc(slope))

```

```{r}
state_slopes %>%
mutate(rank = row_number()) %>%
filter(state == "West Virginia")

```

***Interpretation***

West Virginia exhibits one of the steepest positive gradients in obesity
prevalence among U.S. states. Unlike Florida, this pattern is consistent
and monotonic, suggesting a real and persistent structural risk rather
than random fluctuation.

This result aligns with known socioeconomic and health disparities in
the region, including limited access to preventive healthcare,
structural poverty, and reduced opportunities for physical activity.

## 8. Visualizations

```{r}
ggplot(
filter(brfss_clean, state %in% c("Florida", "West Virginia")),
aes(x = year, y = obesity, color = state)
) +
geom_point() +
geom_smooth(method = "lm", se = FALSE) +
labs(
title = "Obesity Patterns in Florida and West Virginia",
x = "Year",
y = "Obesity Prevalence (%)"
)

```

```{r}
plot_data_clean <- brfss_clean %>%
  filter(state %in% c("Florida", "West Virginia")) %>%
  group_by(state, year) %>%
  summarise(
    avg_obesity = mean(obesity, na.rm = TRUE),
    .groups = "drop"
  )

```

```{r}
ggplot(plot_data_clean, aes(x = year, y = avg_obesity, color = state)) +
  geom_line(size = 1.3) +
  geom_point(size = 2.5) +
  scale_color_manual(
    values = c("Florida" = "#1f4e79", "West Virginia" = "#d2691e")
  ) +
  labs(
    title = "Obesity Trend Comparison: Florida vs West Virginia (2011–2023)",
    x = "Years",
    y = "Average Obesity Rate (%)",
    color = "State"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    plot.title = element_text(face = "bold", hjust = 0.5),
    axis.title = element_text(face = "bold"),
    legend.position = "right"
  )


```

***Figure:*** Average adult obesity prevalence in Florida and West
Virginia by year. Florida exhibits greater year-to-year variability,
while West Virginia maintains a consistently higher and steadily
increasing obesity trajectory.

## 9. Answering the Individual Sub-Question (SQ)

What caused obesity rates in Florida and West Virginia to act “weird”?

Florida and West Virginia appear atypical for fundamentally different
statistical and substantive reasons. Florida’s obesity estimates display
unusually high year-to-year variability, a pattern that is inconsistent
with realistic population-level change in a slow-moving health outcome
such as obesity. This volatility is best explained by survey-related
instability, including changes in sample composition, weighting
procedures, and the state’s large and heterogeneous population, which
can amplify fluctuations in annual estimates. Importantly, Florida’s
long-term average obesity prevalence remains relatively stable,
reinforcing the interpretation that the observed “zig-zag” behavior
reflects measurement noise rather than true behavioral shifts.

In contrast, West Virginia exhibits a strong, persistent, and
directional obesity pattern. Its obesity prevalence is consistently
among the highest in the nation and follows a smooth upward trajectory
across years. This consistency suggests a genuine structural health risk
rather than a statistical artifact. Unlike Florida, West Virginia’s
pattern does not rely on short-term fluctuations but reflects enduring
socioeconomic, demographic, and environmental conditions that contribute
to elevated obesity risk.

Together, these findings illustrate that apparent anomalies in public
health data can arise from either measurement instability or meaningful
population differences. Distinguishing between these mechanisms is
essential for accurate interpretation, as treating both states as
equivalent “outliers” would obscure the true public health challenges
represented by West Virginia while overstating the significance of
Florida’s variability.

## 10. Connection Back to the Group Overall Question (OQ)

This analysis demonstrates that the apparent anomalies observed in
Florida and West Virginia arise from fundamentally different sources.
Florida’s irregular pattern reflects data instability rather than true
behavioral change, while West Virginia represents a genuine geographic
outlier with persistently elevated obesity rates.

By distinguishing between measurement noise and structural trends, this
sub-question strengthens the group’s overall conclusions. It ensures
that state-level deviations are interpreted appropriately, preventing
misleading inferences about national or regional obesity dynamics.

## 11. Limitations

Several limitations should be noted. The BRFSS relies on self-reported
height and weight, which may underestimate true obesity prevalence.
Additionally, annual estimates are subject to survey design changes,
nonresponse bias, and sampling variability, particularly in large or
demographically diverse states such as Florida.

These limitations reinforce the importance of cautious interpretation,
especially when evaluating short-term fluctuations rather than long-term
trends.

## 12. Conclusion

This individual analysis investigated why obesity trends in Florida and
West Virginia appeared atypical during the group’s exploratory analysis.
Although both states initially stood out as anomalies, closer
examination revealed that their unusual patterns stem from fundamentally
different sources.

Florida’s apparent “zig-zag” behavior was driven by substantial
year-to-year variability in obesity estimates rather than sustained
directional change. Given that obesity is a slow-moving population
health outcome, such volatility is unlikely to reflect true behavioral
shifts. Instead, Florida’s pattern is best explained by survey-related
factors, including sampling variability, population heterogeneity, and
weighting effects inherent in large, diverse states. As a result,
Florida should not be interpreted as a genuine outlier in obesity
trends.

In contrast, West Virginia exhibited consistently high obesity
prevalence across multiple years, with a steady upward trajectory.
Unlike Florida, this pattern remained stable over time, indicating a
persistent structural obesity burden rather than measurement noise. This
finding suggests that West Virginia represents a true geographic outlier
with long-standing public health challenges that warrant sustained
attention and intervention.

By distinguishing between data instability and meaningful structural
trends, this analysis prevents over-interpretation of noisy survey
estimates and strengthens the validity of the group’s overall
conclusions. More broadly, it highlights the importance of diagnostic
checks when interpreting public health data and demonstrates how
apparent anomalies can reveal either limitations of measurement or
genuine population-level risk.
